From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-22 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Adding changes from meeting

In the continuous case, when $\mathcal{S}$ is a manifold, we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. We will call a set of quantities $\xi^a$ \textit{state variables} provided that they fully identify a state, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates'. However, we will have multiple manifolds and, to avoid confusion, we will call `state variables' the coordinates of the state space.} We define a density $\rho^{\ast}_c$: $\mathbb{R}^n \to \mathbb{R}$ as the pullback of the distribution $\rho_c$ by the inverse of a coordinate map:  $\rho^{\ast}_c \equiv (\rho_c \circ \xi^{-1})$. %To represent the distribution over state variables, we define coordinate charts $(U, \xi)$ over $\mathcal{S}$. Likewise, the distribution $\rho_c$ is the pullback of $\rho^{\ast}_c$ by a coordinate map: $\rho_c (s) = (\xi^{\ast} \rho^{\ast}_c)(s) = \rho^{\ast}_c (\xi (s))$.

The distribution $\rho_c$ is a function of the particle states $s \in \mathcal{S}$, while the density $\rho^{\ast}_c$ is a function of the state variables $\xi^a \in \mathbb{R}^n$. A coordinate map $\xi$ lets us change variables from $s$ to $\xi^a$. Consequently, we require that $\mu(U) = \int_U \rho_c(s) d\mathcal{S} = \int_{\xi(U)} \rho^{\ast}_c (\xi^a) d \xi^a$.\footnote{Indeed, using the Lebesgue measure $\lambda$ on $\mathbb{R}^n$---and assuming that $\xi^{-1}$: $\mathbb{R}^n \to \mathcal{S}$ is measurable---we can define the measure $d\mathcal{S}$ by the pushforward measure $\xi^{-1}_{\ast}(\lambda)(U) \equiv \lambda(\xi(U))$.} Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$. Hence, $\rho_c$ is a scalar function and consequently coordinate independent. Thus, $\int_U \rho_c(s) d\mathcal{S}$ is invariant under coordinate changes $\xi$ to $\hat{\xi}$. As a function of the state variables $\xi^a$, $\rho^{\ast}_c$ transforms according to the Jacobian for this change of variables: $\rho^{\ast}_c (\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c (\hat{\xi}^b)$. Since $\mu(U) = \int_{\xi(U)} \rho^{\ast}_c (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho^{\ast}_c (\hat{\xi}^b) d \hat{\xi}^b$ remains invariant under this coordinate change, the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless.\footnote{Formally, we can express this argument as a chain of equalities: $\rho^{\ast}_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho^{\ast}_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c(\hat\xi^b)$, where $\rho^{\ast}_c(\xi^a)$ and $\rho^{\ast}_c(\hat{\xi}^b)$ are the same density expressed over different state variables and $s(\xi^a) \equiv \xi^{-1}(\xi^a)$ and $s(\hat{\xi}^b)\equiv \hat{\xi}^{-1}(\hat{\xi}^b)$ represent the same particle state as expressed by these different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through their labeling of a particle state $s$.} 
%$\rho^{\ast}_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho^{\ast}_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c(\hat\xi^b)$

@@

The setting is the extended phase space $(q, p, t)$, which is the standard phase space extended by the time variable. Knowing that the system is deterministic and reversible, it follows that the density of states is invariant. This means that possible trajectories of the system in the extended phase space are never pushed toward each other or pulled apart over time. In other words, the flux of trajectories through any closed surface in the extended phase space is always zero. Formally, the vector field $\vec{S} := (\frac{d q }{d t }, \frac{d p }{d t }, \frac{d t }{d t })$ represents the displacement of states along the trajectories. This field is divergenceless, $\nabla \cdot \vector{S} = 0$, because, given a region $U$ as many states will enter the region as exit. In particular, note that the displacement is tangent to the trajectory $\gamma [q, p, t] (t)$, in extended phase space, where the variable $t$ functions as both a coordinate and the parameter of the trajectory. [XXX:would avoid using affine parameter... it may not be affine]

@@

Nonetheless, reformulating Hamiltonian mechanics within the extended phase space $(q_i, p_i, t, E)$ provides one way to rectify these deficiencies. At root, the time-dependence of the Hamiltonian stems from the dual role of time as both a coordinate and an affine parameter within the usual Hamiltonian framework. Yet, differential geometry demands that we separate these two roles. We thereby add a temporal degree of freedom $(t, E) $, expanding the phase space degrees of freedom by one (recall that we previously had one degree of freedom for every canonical coordinate pair $q_i, p_i$). Next, we define the phase space evolution over an affine parameter $s $. This yields two more equations of motion: $dt/ds = -\partial \mathcal{H}/ \partial E$ and $dE/ds = -\partial \mathcal{H}/ \partial t$, where $\mathcal{H}(q_i, p_i, t, E)$ is the extended Hamiltonian. In this framework, there is a \textit{Hamiltonian constraint} $\mathcal{H}=0=H(q_i, p_i, t)-E$ that functions as both a global constraint---fixing the energy as a function of $q, p, t $---and as a generator of the affine parameter $s $ [XXX to do: provide a reference for extened phase space formulation. And check this discussion of the `Hamiltonian constraint']. Crucially, the extended Hamiltonian---unlike the ordinary Hamiltonian---is coordinate-invariant, even with respect to time-dependent coordinate transformations. For example, if we define the affine parameter as the proper time, then the extended Hamiltonian is proportional to the mass (which is manifestly coordinate-invariant). This provides a clear physical meaning for the extended Hamiltonian. We have now considered many conceptual and physical advantages of the Hamiltonian formalism. In the next section, we rebut Curiel's \parencites*[]{Curiel} arguments against privileging Hamiltonian mechanics. 
%%Arnold uses a related extended phase space $(q_i, p_i, t)$ in his chapter 9, but doesn't provide a general discussion of it. 
%Discussion by GC of extended phase space: https://tinyurl.com/4tlbej76
% Is the Hamiltonian constraint for the extended phase space the same as the discussion in this Wikipedia page: https://en.wikipedia.org/wiki/Hamiltonian_constraint

>>>>

In the continuous case, when $\mathcal{S}$ is a manifold, we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. We will call a set of quantities $\xi^a$ \textit{state variables} provided that they fully identify a state, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates'. However, we will have multiple manifolds and, to avoid confusion, we will call `state variables' the coordinates of the state space.} We define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ as the pullback of the distribution $\rho_c$ by the inverse of a coordinate map:  $\rho_{c,\xi} \equiv (\rho_c \circ \xi^{-1})$. %To represent the distribution over state variables, we define coordinate charts $(U, \xi)$ over $\mathcal{S}$. Likewise, the distribution $\rho_c$ is the pullback of $\rho^{\ast}_c$ by a coordinate map: $\rho_c (s) = (\xi^{\ast} \rho^{\ast}_c)(s) = \rho^{\ast}_c (\xi (s))$.

The distribution $\rho_c$ is a function of the particle states $s \in \mathcal{S}$, while the density $\rho_{c, \xi}$ is a function of the state variables $\xi^a \in \mathbb{R}^n$. A coordinate map $\xi$ lets us express the state $s$ as a function of the coordinates $\xi^a$. Consequently, we require that $\mu(U) = \int_U \rho_c(s) d\mathcal{S} = \int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a$.\footnote{Indeed, using the Lebesgue measure $\lambda$ on $\mathbb{R}^n$---and assuming that $\xi^{-1}$: $\mathbb{R}^n \to \mathcal{S}$ is measurable---we can define the measure $d\mathcal{S}$ by the pushforward measure $\xi^{-1}_{\ast}(\lambda)(U) \equiv \lambda(\xi(U))$.} Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$. Hence, $\rho_{c, \xi}$ will transform as a scalar under change of variables. Thus, $\int_U \rho_c(s) d\mathcal{S}$ is invariant under coordinate changes $\xi$ to $\hat{\xi}$. As a function of the state variables $\xi^a$, $\rho^{\ast}_c$ transforms according to the Jacobian for this change of variables: $\rho^{\ast}_c (\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c (\hat{\xi}^b)$. Since $\mu(U) = \int_{\xi(U)} \rho^{\ast}_c (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho^{\ast}_c (\hat{\xi}^b) d \hat{\xi}^b$ remains invariant under this coordinate change, the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless.\footnote{Formally, we can express this argument as a chain of equalities: $\rho^{\ast}_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho^{\ast}_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c(\hat\xi^b)$, where $\rho^{\ast}_c(\xi^a)$ and $\rho^{\ast}_c(\hat{\xi}^b)$ are the same density expressed over different state variables and $s(\xi^a) \equiv \xi^{-1}(\xi^a)$ and $s(\hat{\xi}^b)\equiv \hat{\xi}^{-1}(\hat{\xi}^b)$ represent the same particle state as expressed by these different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through their labeling of a particle state $s$.} 
%$\rho^{\ast}_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho^{\ast}_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c(\hat\xi^b)$

@@

The setting is the extended phase space $(q, p, t)$, which is the standard phase space extended by the time variable. Knowing that the system is deterministic and reversible, it follows that the density of states is invariant. This means that possible trajectories of the system in the extended phase space are never pushed toward each other or pulled apart over time. In other words, the flux of trajectories through any closed surface in the extended phase space is always zero. Formally, the vector field $\vec{S} := (\frac{d q }{d t }, \frac{d p }{d t }, \frac{d t }{d t })$ represents the displacement of states along the trajectories. This field is divergenceless, $\nabla \cdot \vector{S} = 0$, because, given a region $U$ as many states will enter the region as exit. In particular, note that the displacement is tangent to the trajectory $\gamma [q, p, t] (t)$, in extended phase space, where the variable $t$ functions as both a coordinate and the parameter of the trajectory.

@@

Nonetheless, reformulating Hamiltonian mechanics within the extended phase space $(q_i, p_i, t, E)$ provides one way to rectify these deficiencies. At root, the time-dependence of the Hamiltonian stems from the dual role of time as both a coordinate and a time-like parameter within the usual Hamiltonian framework. Yet, differential geometry demands that we separate these two roles. We thereby add a temporal degree of freedom $(t, E) $, expanding the phase space degrees of freedom by one (recall that we previously had one degree of freedom for every canonical coordinate pair $q_i, p_i$). Next, we define the phase space evolution over an evolution parameter $s $. This yields two more equations of motion: $dt/ds = -\partial \mathcal{H}/ \partial E$ and $dE/ds = -\partial \mathcal{H}/ \partial t$, where $\mathcal{H}(q_i, p_i, t, E)$ is the extended Hamiltonian. The extended Hamiltonian not only acts as a generator of the evolution parameter $s$, but it also provides us with the \textit{Hamiltonian constraint} $\mathcal{H}=0$ [XXX to do: provide a reference for extened phase space formulation. And check this discussion of the `Hamiltonian constraint']. The Hamiltonian constraint provides the relationship between the energy and the other state variables. Crucially, the extended Hamiltonian---unlike the ordinary Hamiltonian---is coordinate-invariant, even with respect to time-dependent coordinate transformations. For example, if we define the affine parameter as the proper time, then the extended Hamiltonian is proportional to the mass (which is manifestly coordinate-invariant). This provides a clear physical meaning for the extended Hamiltonian. We have now considered many conceptual and physical advantages of the Hamiltonian formalism. In the next section, we rebut Curiel's \parencites*[]{Curiel} arguments against privileging Hamiltonian mechanics. 
%%Arnold uses a related extended phase space $(q_i, p_i, t)$ in his chapter 9, but doesn't provide a general discussion of it. 
%Discussion by GC of extended phase space: https://tinyurl.com/4tlbej76
% Is the Hamiltonian constraint for the extended phase space the same as the discussion in this Wikipedia page: https://en.wikipedia.org/wiki/Hamiltonian_constraint
